Convolutional Neural Nets for the MNIST Digit Dataset

### 2014-12-13 tex/paper.tex
proofread paper

%%%%%%%%% ABSTRACT
\begin{abstract}
   In this paper we describe the design and implementation of a convolutional neural net (a CNN)
   that we use to classify digits from the popular MNIST dataset. We first lay a foundation
   for the merit of convolutional neural nets in object recognition, and then build on this
   foundation to explain our design decisions. The resulting CNN is able to achieve 94\% on the
   MNIST digit dataset while training on only 2\% of the MNIST training set. We feel that the accuracy
   of our CNN could still be improved by training on more images (especially deformed versions of images).
\end{abstract}

@@

Convolutional neural nets recognize this difficulty and tackle it by iteratively tuning randomly
initialized filters (that act on all locations of an image) to produce a well-known output using
the back-propagation algorithm. These neural nets are convolutional because the convolution operator
is used to effectively share filters across an entire image. While time-consuming to train, CNN's
can provide world-class performance on a large and diverse dataset \cite{ImageNet}.
To gain a deeper understanding of the challenges of implementing an object recognition system,
we designed and implemented a convolutional neural net and tested it on the popular MNIST
digit dataset \cite{MNIST}.
To examine the effect of the number of layers in convolutional neural networks,
we trained two versions of our net with different depths on the MNIST training data
and compare their performance here.

@@

We can construct a different way to use a hierarchical representation of an image for classification
by performing convolutions at different scales of images with sets of filters.
Convolutional Neural Nets (CNNs) use a combination of convolutional layers, which perform convolution
of an image with a set of filters, and pooling layers, which reduce the dimensionality of the output
of convolutional layers, to create this representation. Convolutional layers serve to share the filter weights
across the image and thus reduce the number of parameters that must be trained, while also providing
a certain degree of traslational equivariance.
Pooling layers reduce dimensionality by taking the maximum value (max-pooling) or mean value
(mean-pooling) of a region of convolutional outputs.
This reduction of dimensionality is important because it helps reduce the number of parameters that
must be trained in the network, thus speeding up learning.

@@

The training of CDBNs is accomplished using the same greedy approach as in DBNs.
Once the parameters have been learned, the CDBNs representation of an image can be computed
by sampling from the joint distribution over all the hidden layers given the input image.
The classification of images can be produced by training a classification algorithm
on the higher level features.

@@

\begin{figure*}
  \includegraphics[width=\textwidth,height=8cm]{convtrans}
  \caption{Given an input image $I$ of size $w_{img} \times w_{img}$ with $c$ channels and $K$ kernels of size
  $w_k \times w_k$ with $c$ channels, each region on which the kernel is applied can be flattened to form a
  row of a matrix $I’$ of size $(w_{img} - w_k + 1)^2 \times c \cdot w_k^2$. The set of kernels can also be
  flattened to form columns of a matrix $W$ of size $c \cdot w_k^2 \times K$. If we take the matrix product
  of $I’$ and $W$, then the $i^\text{th}$ column of the resultant matrix is a flattened version of the image
  convolved with the $i^\text{th}$ kernel. The matrix $W$ can be seen as the weights of a fully connected
  network with $c \cdot w_k^2$ inputs and $K$ outputs. Thus, learning of the kernels is reduced to batch
  learning of the weights of this fully connected network.
}
  \label{fig:convtrans}
\end{figure*}

@@

The convolutional layer learns a set of small filters that when convolved with the input image
during forward propagation extract salient features. This is done through the use of local
connectivity and parameter sharing. Local connectivity refers to each unit of the convolutional
layer being connected only to a small region of the image; this region is known as the unit’s
receptive field, and the weights of these connections form a filter. Neighboring units in the
convolutional layer have overlapping receptive fields, with each receptive field being shifted
by one pixel as you move along a certain dimension. These units are then organized into feature
maps that share the same weights on their connections. This combination of local connectivity
and parameter sharing results in a network with sparser connectivity and fewer parameters than
would be present in a fully connected network.

Though one could implement the convolutional layer directly by building the sparse network and
applying the appropriate weight updates during the backpropagation step, it is also possible to
transform this operation into the form of batch learning of the weights of a fully connected
network (discussion in section \ref{sec:full}), such that

$$ I' \times W = [I*k_1 \hspace{10pt} I*k_2 \hspace{10pt} \cdots \hspace{10pt} I*k_K]$$
where $I$ is the image, $W$ is the weight matrix of the fully connected network, $I$’ is the transformed
image, and $k_i$ is the $i^{\text{th}}$ kernel. The learned filters would then correspond to slices of the kernels.
We took the later approach for ease of implementation. Figure \ref{fig:convtrans} demonstrates how this transformation
is accomplished.

@@

The pooling layer reduces the dimension of its input by a factor of n. During forward propagation, each
unit in the pooling layer looks at an n x n window and applies some pooling operator, either max or mean.
Though these n x n windows could in principle overlap, in our implementation they do not. No learning occurs
in this layer, thus backpropagation only serves to directly forward the errors to the previous layer.
If the pooling layer did max-pooling, then each pooling unit propagates the error to the input unit that
had the max value with all other errors in that window being zero. If the pooling layer did mean-pooling,
then each pooling unit propagates the value of the error divided by the number of input units is passed to all input units.

@@

Because a single input image is broken up into multiple samples during convolution, our
fully connected layer is capable of batch training, where the update to a set of weights over
many samples is the sum of updates for each sample. While batch training as a whole is generally
slower than on-line training \cite{EfficientBackProp}, we are still performing on-line training
with respect to each input image, so we do not take an unreasonable performance hit.

@@

Another way we improved the likelihood of convergence was to make each weight update of a
fully connected layer also depend on the previous weight update (this is sometimes
referred to as momentum). With this change, the weight update equation for a fully connected
layer becomes
$$\Delta W_{t} = W_{t} - (\eta \frac{\partial E}{\partial W_{t}} +
\alpha \Delta W_{t-1})$$
where $\eta$ is the learning rate and $\alpha$ is the momentum.
Because many updates to $W$ have the same sign, this can effective speed up training time
\cite{WinMomentum}.
Finding good values for the momentum and learning rate seems to be
an exercise in trial-and-error, although we did find that
the learning rates we used tended to be below $0.2$.

One important suggestion from \cite{EfficientBackProp} that we did not use was the
recommended sigmoid non-linear output function of $f(x) = 1.7159 * tanh(\frac{2}{3}x)$.
While we experimented with the standard sigmoid of $f(x) = \frac{1}{1 + exp(-x)}$,
we used $f(x) = tanh(x)$ for most of our results.

@@

The following experiments were conducted using the MNIST dataset \cite{MNIST}, which includes a training set of
60000 28x28 handwritten digit images and a test set of 10,000 28x28 handwritten digit images.
It is important to note that both the training and test sets are composed of two sets of digits:
an easier set written by US Census Bureau employees and a harder set written by high school students.

@@

The one convolutional/pooling layer architecture was trained for 50 iterations with 50 images per digit
(training set of 500 images), and was subsequently tested on
1,000 new images per digit (test set of 10,000 images).
The training images were taken from the easier portion of the training set, while the test images were taken from
both sets. The results of this training can be seen in Figure \ref{fig:accplot}. This trained network achieved an
accuracy of 94\% and very quickly converged after five iterations of training. An example
classification of an image with the activations and learned filters is presented
in Figure \ref{fig:actfilts}. Figure \ref{fig:misclassify} shows examples of correctly and incorrectly classified images.

\begin{figure}
  \includegraphics[scale=0.55]{AccPlot}
  \caption{Performance of differnet architectures and training set size. The one convolutional/pooling architecture with a training set
  of size 500 achieved an accuracy of 94\%, while it achieved an accuracy of 93.22\% on a training set of size 1000. The two convolutional/pooling
  architecture achieved an accuracy of 92.8\%. Accuracy testing was performed after every five iterations of training.}
  \label{fig:accplot}
\end{figure}

@@

The second experiment with the one convolutional/pooling layer architecture was also trained
for 50 iterations but with 100 images per digit instead of 50 (training set of 1,000 images).
This trained network was then tested on 1,000 new images per digit (test set of 10,000 images).
The results of this training can be seen in Figure \ref{fig:accplot}.This trained network
achieved an accuracy of 93.22\% and also quickly converged after about five iterations of training.

@@

The two convolutional/pooling layer architecture was also trained for 50 iterations with 50 images per digit
(training set of 500 images), and was subsequently tested on 1,000 new images per digit
(test set of 10,000 images).
The training images were taken from the easier portion of the training set,
while the test images were taken from both sets.
The results of this training can be seen in Figure \ref{fig:accplot}. This trained network achieved an accuracy of 92.8\% converging after about 10 iterations of training.
An example classification of an image with the activations and learned filters is presented
in Figure \ref{fig:actfilts}.

@@

\begin{figure*}
  \makebox[\textwidth][c]{\includegraphics[scale=0.72]{misclassify}}
  \caption{This is a tiger. Rawr!}
  \label{fig:misclassify}
\end{figure*}

>>>>

%%%%%%%%% ABSTRACT
\begin{abstract}
   In this paper we describe the design and implementation of a convolutional neural net (a CNN)
   that we use to classify digits from the popular MNIST dataset. We first lay a foundation
   for the merit of convolutional neural nets in object recognition, and then build on this
   foundation to explain our design decisions. The resulting CNN is able to achieve 94\% on the
   MNIST digit test set while training on only 2\% of the MNIST training set. We feel that the accuracy
   of our CNN could still be improved by training on more images (especially deformed versions of images).
\end{abstract}

@@

Convolutional neural nets recognize this difficulty and tackle it by iteratively tuning randomly
initialized filters (that act on all locations of an image) to produce a well-known output using
the back-propagation algorithm. These neural nets are convolutional because the convolution operator
is used to effectively share filters across an entire image. While time-consuming to train, CNN's
can provide world-class performance on a large and diverse dataset \cite{ImageNet}.
To gain a deeper understanding of the challenges of implementing an object recognition system,
we designed and implemented a convolutional neural net and tested it on the popular MNIST
digit dataset \cite{MNIST}.
To examine the effect of the number of layers in convolutional neural networks,
we trained two versions of our net of different depths on the MNIST training data
and compare their performance here.

@@

We can construct a different way to use a hierarchical representation of an image for classification
by performing convolutions at different scales of images with sets of filters.
Convolutional Neural Networks (CNNs) use a combination of convolutional layers, which perform convolution
of an image with a set of filters, and pooling layers, which reduce the dimensionality of the output
of convolutional layers, to create this representation. Convolutional layers serve to share the filter weights
across the image and thus reduce the number of parameters that must be trained, while also providing
a certain degree of translational equivariance.
Pooling layers reduce dimensionality by taking the maximum value (max-pooling) or mean value
(mean-pooling) of a region of convolutional outputs.
This reduction of dimensionality is important because it helps reduce the number of parameters that
must be trained in the network, thus speeding up learning.

@@

The training of CDBNs is accomplished using the same greedy approach as in DBNs.
Once the parameters have been learned, the CDBN's representation of an image can be computed
by sampling from the joint distribution over all the hidden layers given the input image.
The classification of images can be produced by training a classification algorithm
on the higher level features.

@@

\begin{figure*}
  \includegraphics[width=\textwidth,height=8cm]{convtrans}
  \caption{Given an input image $I$ of size $w_{img} \times w_{img}$ with $c$ channels and $K$ kernels of size
  $w_k \times w_k$ with $c$ channels, each region on which the kernel is applied can be flattened to form a
  row of a matrix $I^{\prime}$ of size $(w_{img} - w_k + 1)^2 \times c \cdot w_k^2$. The set of kernels can also be
  flattened to form columns of a matrix $W$ of size $c \cdot w_k^2 \times K$. If we take the matrix product
  of $I^{\prime}$ and $W$, then the $i^\text{th}$ column of the resultant matrix is a flattened version of the image
  convolved with the $i^\text{th}$ kernel. The matrix $W$ can be seen as the weights of a fully connected
  network with $c \cdot w_k^2$ inputs and $K$ outputs. Thus, learning of the kernels is reduced to batch
  learning of the weights of this fully connected network.
}
  \label{fig:convtrans}
\end{figure*}

@@

The convolutional layer learns a set of small filters that when convolved with the input image
during forward propagation extract salient features. This is done through the use of local
connectivity and parameter sharing. Local connectivity refers to each unit of the convolutional
layer being connected only to a small region of the image; this region is known as the unit's
receptive field, and the weights of these connections form a filter. Neighboring units in the
convolutional layer have overlapping receptive fields, with each receptive field being shifted
by one pixel as you move along a certain dimension. These units are then organized into feature
maps that share the same weights on their connections. This combination of local connectivity
and parameter sharing results in a network with sparser connectivity and fewer parameters than
would be present in a fully connected network.

Though one could implement the convolutional layer directly by building the sparse network and
applying the appropriate weight updates during the back-propagation step, it is also possible to
transform this operation into the form of batch learning of the weights of a fully connected
network (discussion in section \ref{sec:full}), such that

$$ I' \times W = [I*k_1 \hspace{10pt} I*k_2 \hspace{10pt} \cdots \hspace{10pt} I*k_K]$$
where $I$ is the image, $W$ is the weight matrix of the fully connected network, $I'$ is the transformed
image, and $k_i$ is the $i^{\text{th}}$ kernel. The learned filters would then correspond to slices of the kernels.
We took the later approach for ease of implementation. Figure \ref{fig:convtrans} demonstrates how this transformation
is accomplished.

@@

The pooling layer reduces the dimension of its input by a factor of $n$. During forward propagation, each
unit in the pooling layer looks at an $n$ x $n$ window and applies some pooling operator, either max or mean.
Though these $n$ x $n$ windows could in principle overlap, in our implementation they do not. No learning occurs
in this layer, thus back-propagation only serves to directly forward the errors to the previous layer.
If the pooling layer did max-pooling, then each pooling unit propagates the error to the input unit that
had the max value with all other errors in that window being zero. If the pooling layer does mean-pooling,
then each pooling unit propagates the value of the error divided by the number of input units to all input units.

@@

Because a single input image is broken up into multiple samples during convolution, our
fully connected layer is capable of batch training, where the update to a set of weights over
many samples is the sum of updates for each sample. While batch training as a whole is generally
slower than on-line training \cite{EfficientBackProp}, we are still performing on-line training
with respect to the set of input images, so we do not take an unreasonable performance hit.

@@

Another way we improved the likelihood of convergence was to make each weight update of a
fully connected layer also depend on the previous weight update (this is sometimes
referred to as momentum). With this change, the weight update equation for a fully connected
layer becomes
$$\Delta W_{t} = -(\eta \frac{\partial E}{\partial W_{t}} +
\alpha \Delta W_{t-1})$$
where $\eta$ is the learning rate and $\alpha$ is the momentum.
Because many updates to $W$ have the same sign, this can effective speed up training time
\cite{WinMomentum}.
Finding good values for the momentum and learning rate seems to be
an exercise in trial-and-error, although we did find that
the learning rates we used tended to be below $0.2$.

One important suggestion from \cite{EfficientBackProp} that we did not use was the
recommended sigmoid non-linear output function of $f(x) = 1.7159 * tanh(\frac{2}{3}x)$.
While we experimented with the standard sigmoid of $f(x) = \frac{1}{1 + exp(-x)}$,
we found $f(x) = tanh(x)$ to perform best for most of our experiments.

@@

The following experiments were conducted using the MNIST dataset \cite{MNIST}, which includes a training set of
60,000 28x28 handwritten digit images and a test set of 10,000 28x28 handwritten digit images.
It is important to note that both the training and test sets are composed of two sets of digits:
an easier set written by US Census Bureau employees and a harder set written by high school students.

@@

Our one convolutional/pooling layer architecture was trained for 50 iterations with 50 images per digit
(for a training set of 500 images), and was subsequently tested on 1,000 new images per digit
(for a test set of 10,000 images).
The training images were taken from the easier portion of the training set, while the test images were taken from
both sets. The results of this training can be seen in Figure \ref{fig:accplot}. This trained network achieved an
accuracy of 94\% and very quickly converged after five iterations of training. An example
classification of an image with the activations and learned filters is presented
in Figure \ref{fig:actfilts}. Figure \ref{fig:misclassify} shows examples of correctly and incorrectly classified images on this architecture.

\begin{figure}
  \includegraphics[scale=0.55]{AccPlot}
  \caption{Performance of our different architectures labeled by training set size (all trained for 50
  iterations).
  The one convolutional/pooling architecture with a training set of 500 images achieved an accuracy of 94\%,
  while with a training set of 1,000 images it achieved an accuracy of 93.22\%.
  The two convolutional/pooling architecture achieved an accuracy of 92.8\%.
  Accuracy testing was performed after every five iterations of training.}
  \label{fig:accplot}
\end{figure}

@@

The second experiment with our one convolutional/pooling layer architecture was also trained
for 50 iterations but with 100 images per digit instead of 50 (for a training set of 1,000 images).
This trained network was then tested on 1,000 new images per digit (for a test set of 10,000 images).
The results of this training can be seen in Figure \ref{fig:accplot}. This trained network
achieved an accuracy of 93.22\% and also quickly converged after about five iterations of training.

@@

The two convolutional/pooling layer architecture was also trained for 50 iterations with 50 images per digit
(for a training set of 500 images), and was subsequently tested on 1,000 new images per digit
(for a test set of 10,000 images).
The training images were taken from the easier portion of the training set,
while the test images were taken from both sets.
The results of this training can be seen in Figure \ref{fig:accplot}. This trained network achieved an accuracy of 92.8\% converging after about 10 iterations of training.
An example classification of an image with the activations and learned filters is presented
in Figure \ref{fig:actfilts}.

@@

\begin{figure*}
  \makebox[\textwidth][c]{\includegraphics[scale=0.72]{misclassify}}
  \caption{The left side shows examples of correctly classified digits by the one convolutional/pooling layer architecture, while the right side shows examples of incorrectly
  classified digits by the same architecture. The top right corner shows the correct label. The numbers below the image are the first and second place labels determined by our
  CNN. Note how the misclassified images can be seen to resemble the network's chosen label. The number 8 was misclassified multiple times, likely due to its similarity to other
  digits when deformed from its canonical form.}
  \label{fig:misclassify}
\end{figure*}
